\chapter*{Acknowledgements}


\begin{flushright}
    \emph{Living off borrowed time, the clock tick faster \\
    That'd be the hour they knock the slick blaster} \\
    \textsf{-- MF DOOM}
\end{flushright}

\vspace*{1cm}

This thesis represent the end of almost ten years living in Bologna and studying Physics, which I can say without a doubt that have been the best years of my life so far.

\medskip

First and foremost, I would like to thank my supervisor Elisa Ercolessi, for following and supporting me since my Master.
If it was not for her courses on Statistical Mechanics and Quantum Many Body Theory, I would not have been inspired in doing Theoretical Physics.

\smallskip

Then, I would like to thank my parents and my family for their unconditional support of my studies, even though they still, to this day, have not a clear idea of what I do.
Nonetheless, I am certain that this career choice of mine makes them proud.

\smallskip

Next, I would like to thank all the friends I have made over the years.
From the classmates and office colleagues to the roommates, without forgetting all the people I have met outside the department.
It is still astonishing to me how many relationships and close friends I have been able to build over the years, which feels like a blessing.
They are the main reason I can rest assured these have been the best years of my life so far.

\smallskip

Last but not least, a special thanks to the jungle music scene of the 90s, and record labels like Moving Shadow, Good Looking Records, Suburban Base Records, etc.
Without their sound, the writing of this thesis would have not been possible.
